---
title: Problem set 4
date: 2025-10-05
---

In the next problem set, we plan to explore the relationship between COVID-19 death rates and vaccination rates across US states by visually examining their correlation. This analysis will involve gathering COVID-19 related data from the CDC's API and then extensively processing it to merge the various datasets. Since the population sizes of states vary significantly, we will focus on comparing rates rather than absolute numbers. To facilitate this, we will also source population data from the US Census to accurately calculate these rates. 

In this problem set we will learn how to extract and wrangle data from the data US Census and CDC APIs.

1. Get an API key from the US Census at <https://api.census.gov/data/key_signup.html>. You can't share this public key. But your code has to run on a TFs computer. Assume the TF will have a file in their working directory named `census-key.R` with the following one line of code:

```
census_key <- "A_CENSUS_KEY_THAT_WORKS"
```

Write a first line of code for your problem set that defines `census_key` by running the code in the file `census-key.R`.

```{r}
# stored the key in census-key.R so it isn't hard-coded here
census_key <- source("census-key.R")
```



census_key <- "e5d44a807812c3098c41692f3da8f37cf84e7082"



2. The [US Census API User Guide](https://www.census.gov/content/dam/Census/data/developers/api-user-guide/api-user-guide.pdf)
provides details on how to leverage this valuable resource. We are interested in vintage population estimates for years 2021 and 2022. From the documentation we find that the _endpoint_ is:


```{r}
url <- "https://api.census.gov/data/2021/pep/population"
```

Use the **httr2** package to construct the following GET request.

```
https://api.census.gov/data/2021/pep/population?get=POP_2020,POP_2021,NAME&for=state:*&key=YOURKEYHERE
```

Create an object called `request` of class `httr2_request` with this URL as an endpoint.
Hint: Print out `request` to check that the URL matches what we want.

```{r}
#| message: false
#| warning: false
library(httr2)
request <- request(url) |> req_url_query(
    get = "POP_2020,POP_2021,NAME",
    `for` = "state:*",
    key = census_key$value)
```


3. Make a request to the US Census API using the `request` object. Save the response to and object named `response`. Check the response status of your request and make sure it was successful. You can learn about _status codes_ [here](https://en.wikipedia.org/wiki/List_of_HTTP_status_codes).


```{r}
# send request to Census API and store result in response
response <- request |> req_perform()
# check http status
resp_status(response)

```

4. Use a function from the **httr2** package to determine the content type of your response.

```{r}
resp_content_type(response)
```

5. Use just one line of code and one function to extract the data into a matrix. 
Hints: 1) Use the `resp_body_json` function. 2) The first row of the matrix will be the variable names and this OK as we will fix in the next exercise.

```{r}
population <- resp_body_json(response, simplifyMatrix = TRUE)
```

6. Examine the `population` matrix you just created. Notice that 1) it is not tidy, 2) the column types are not what we want, and 3) the first row is a header. Convert `population` to a tidy dataset. Remove the state ID column and change the name of the column with state names to `state_name`. Add a column with state abbreviations called `state`. Make sure you assign the abbreviations for DC and PR correctly. Hint: Use the **janitor** package to make the first row the header. 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
#population <- population |> ## Use janitor row to names function
  # convert to tibble
  # remove stat column
  # rename state column to state_name
  # use pivot_longer to tidy
  # remove POP_ from year
  # parese all relevant colunns to numeric
  # add state abbreviations using state.abb variable
  # use case_when to add abbreviations for DC and PR

population <- resp_body_json(response) |> 
  do.call(rbind, args = _) |>              # bind list of rows into matrix
  as.data.frame(stringsAsFactors = FALSE) |> 
  row_to_names(row_number = 1) |>           # first row = column names
  as_tibble() |> 
  select(-state) |>                      
  rename(state_name = NAME) |>       
  pivot_longer(
    cols = starts_with("POP_"),
    names_to = "year",
    values_to = "population"
  ) |> 
  mutate(
    year = str_remove(year, "POP_"),       
    year = as.integer(year),
    population = as.numeric(population),
    state = case_when(                     
      state_name == "District of Columbia" ~ "DC",
      state_name == "Puerto Rico" ~ "PR",
      TRUE ~ state.abb[match(state_name, state.name)]
    )
  )
```

7. As a check, make a barplot of states' 2021 and 2022 populations. Show the state names in the y-axis ordered by population size.  Hint: You will need to use `reorder` and use `facet_wrap`.

```{r}
library(ggplot2)
library(tidytext)
# population |> 
  # reorder state
  # assign aesthetic mapping
  # use geom_col to plot barplot
  # flip coordinates
  # facet by year
population |>
  ggplot(aes(x = reorder_within(state_name, population, year),
             y = population)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ year, scales = "free_y") +
  scale_x_reordered() +
  labs(
    x = "State",
    y = "Population",
    title = "State Populations in 2020 and 2021"
  )
```


8. The following URL:

```{r}
url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"
```

points to a JSON file that lists the states in the 10 Public Health Service (PHS) defined by CDC. We want to add these regions to the `population` dataset. To facilitate this create a data frame called `regions` that has two columns `state_name`, `region`, `region_name`. One of the regions has a long name. Change it to something shorter.

```{r}
#| message: false
#| warning: false
library(jsonlite)
library(purrr)
library(tibble)
library(dplyr)
url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"
# regions <- use jsonlit JSON parser 
# regions <- convert list to data frame. You can use map_df in purrr package 

# parse JSON into a list
regions_list <- fromJSON(url, simplifyVector = FALSE)

# flatten the list
regions <- map_df(regions_list, function(x) {
  tibble(
    state_name  = x$states,
    region      = x$region,
    region_name = x$region_name
  )
})

# shorten the region name
regions <- regions |>
  mutate(
    region_name = case_when(
      str_length(region_name) > 30 ~ str_remove(region_name, " - Colorado.*"),
      TRUE ~ region_name
    )
  )
```

9. Add a region and region name columns to the `population` data frame.

```{r}
population <- population |>
  left_join(regions, by = "state_name")
```


10. From reading <https://data.cdc.gov/> we learn the endpoint `https://data.cdc.gov/resource/pwn4-m3yp.json` provides state level data from SARS-COV2 cases. Use the **httr2** tools you have learned to download this into a data frame. Is all the data there? If not, comment on why.


```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"
# make the request
response <- request(api) |> req_perform()

# parse json directly into a df
cases_raw <- resp_body_json(response, simplifyVector = TRUE) |> as_tibble()

# check structure
glimpse(cases_raw)

```
No, all the data is not there. We see 1000 rows because CDC has a default limit, which we can change by adding `$limit=10000000000`.


We see exactly 1,000 rows. We should be seeing over $52 \times 3$ rows per state.

11. The reason you see exactly 1,000 rows is because CDC has a default limit. You can change this limit by adding `$limit=10000000000` to the request. Rewrite the previous request to ensure that you receive all the data. Then wrangle the resulting data frame to produce a data frame with columns `state`, `date` (should be the end date) and `cases`. Make sure the cases are numeric and the dates are in `Date` ISO-8601 format.

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"

# make request with larger limit
response <- request(api) |>
  req_url_query(`$limit` = 10000000000) |>
  req_perform()

# parse json into tibble
cases_raw <- resp_body_json(response, simplifyVector = TRUE) |> as_tibble()

# wrangle into format described
cases <- cases_raw |>
  transmute(
    state = state,
    date = as.Date(end_date),  # Convert ISO-8601 string to date
    cases = as.numeric(tot_cases) # make sure cases are numeric
  )
```

```{r}
is.numeric(population$year)
is.numeric(cases$year)
```


12. For 2020 and 2021, make a time series plot of cases per 100,000 versus time for each state. Stratify the plot by region name. Make sure to label you graph appropriately. 

```{r}
cases_rate <- cases |> 
  left_join(
    population |> select(state, year, population, region_name), # selecting cols we need
    by = c("state", "year") # matching on state/year
  ) |> 
  mutate(cases_per100k = (cases / population) * 100000) |> # calculated cases per 100,000
  filter(year %in% c(2020, 2021)) # for 2020 and 2021


cases_rate |> 
  ggplot(aes(x = date, y = cases_per100k, group = state, color = state)) +
  geom_line() + # line plots for each state
  facet_wrap(~ region_name, scales = "free_y") + #splitting by region
  labs(
    title = "COVID-19 Cases per 100,000 For Each State (2020 - 2021)",
    x = "Date",
    y = "Cases per 100,000",
    color = "State"
  ) +
  theme_minimal()


```

13. The dates in the `cases` dataset are stored as character strings. Use the **lubridate** package to properly parse the `date` column, then create a summary table showing the total COVID-19 cases by month and year for 2020 and 2021. The table should have columns for year, month (as month name), and total cases across all states. Order by year and month. Use the **knitr** package and `kable()` function to display the results as a formatted table.

```{r}
library(knitr)
cases |> 
  # parse character dates into date objects
  mutate(date = ymd(date)) |> 
  
  # extract year and month
  mutate(
    year = year(date),
    month = month(date, label = TRUE, abbr = FALSE) # full month name
  ) |> 
  
  # keeping 2020 and 2021
  filter(year %in% c(2020, 2021)) |> 
  
  # group by year and month
  group_by(year, month) |> 
  
  # total across states
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop") |> 
  
  # order by year then month
  arrange(year, month) |> 
  
  # table
  kable(col.names = c("Year", "Month", "Total Cases"),
        caption = "Monthly COVID-19 Cases Across All States (2020–2021)")
```

14. The following URL provides additional COVID-19 data from the CDC in JSON format:

```{r}
deaths_url <- "https://data.cdc.gov/resource/9bhg-hcku.json"
```

Use **httr2** to download COVID-19 death data from this endpoint. Make sure to remove the default limit to get all available data. Create a clean dataset called `deaths` with columns `state`, `date`, and `deaths` (renamed from the original column name). Ensure dates are in proper Date format and deaths are numeric.


```{r}
# remove default limit
response <- request(deaths_url) |>
  req_url_query(`$limit` = 100000000) |> 
  req_perform()

# parse JSON into tibble
deaths_raw <- resp_body_json(response, simplifyVector = TRUE) |> as_tibble()

# wrangle into clean dataset
deaths <- deaths_raw |> 
  transmute(
    state = state,
    date = as.Date(end_date),         # convert to type date
    deaths = as.numeric(total_deaths)    # make sure deaths are numeric
  ) |> 
  filter(!is.na(date), !is.na(deaths))  # drop empty rows
```

15. Using the `deaths` dataset you created, make a bar plot showing the total COVID-19 deaths by state. Show only the top 10 states with the highest death counts. Order the bars from highest to lowest and use appropriate labels and title.

```{r}
deaths |> 
  # aggregate total deaths across all time for each state
  group_by(state) |> 
  summarise(total_deaths = sum(deaths, na.rm = TRUE), .groups = "drop") |> 
  
  # keep only top 10 states with highest death counts
  slice_max(order_by = total_deaths, n = 10) |> 
  
  ggplot(aes(x = reorder(state, total_deaths), y = total_deaths)) +
  geom_col(fill = "magenta") +
  coord_flip() +  # horizontal bars so you can read the names
  labs(
    title = "Top 10 States by Total COVID-19 Deaths",
    x = "State",
    y = "Total Deaths"
  ) +
  theme_minimal()
```